{\fontfamily{ptm}\selectfont Nesterov Type Momentum Methods}

### 2024-07-08 Notebook LaTEX/Momentum Accelerations/main.tex
>w<, updated notes

    \par\noindent
    \textbf{We state our contributions}. 
    In \hyperref[sec:AG_varieties]{section \ref*{sec:AG_varieties}} 
    We stated different forms of accelerated gradient that appeared in the literatures and the equivalences between their forms. 
    They are abstract because the choice of stepsize parameters in the algorithms are unspecified. 
    In \hyperref[sec:AG_useful_information]{section \ref*{sec:AG_useful_information}}
    we prove and interpret some of the unproved claims made by Ahn, Sra \cite{ahn_understanding_2022} as good exercises and for the peace of the mind.
    \hyperref[lemma:smooth_agg_lyapunov_upper_bound]{Lemma \ref*{lemma:smooth_agg_lyapunov_upper_bound}} 
    rephrase content in the appendix in work by Ahn, Sra\cite{ahn_understanding_2022}. 
    It is our attempt at rephrasing the proofs to create a layer of abstraction so it can be reused for multiple proofs for several variants.

@@

            \begin{definition}[Similar Triangle Form I]
            \label{def:ag_form_similar_tria_I}
                It updates iterates $(y_t, x_{t + 1}, z_{t + 1})$ using 
                \begin{align*}
                    z_{t + 1} &= y_t - L^{-1} \mathcal G_L(y_t) 
                    \\
                    x_{t + 1} &= z_{t + 1} + L\eta_t (z_{t + 1} - z_t)
                    \\
                    y_{t + 1} &= 
                    (1 + L\eta_{t + 1})^{-1}
                    (
                    x_{t + 1} + L\eta_{t + 1}z_{t + 1}
                    ). 
                \end{align*}
            \end{definition}
            \begin{observation}
                There are a lot of observations about this form to unpack. 
                Let's list the following observations
                \begin{enumerate}
                    \item Observe that $y_t = (1 + L\eta_t)^{-1}(x_t + L\eta_t z_t)$ is a convex combination between $x_t, z_t$. Hence $z_t, y_t, x_t$ are three collinear points. Additionally, we have $y_t - x_t = L\eta_t (z_t - y_t)$, hence $\Vert y_t - x_t\Vert/\Vert z_t - y_t\Vert = L\eta_t$. 
                    \item By the algorithm, we have $z_{t + 1} - y_t = - L^{-1} \nabla f(y_t)$, and $x_{t + 1} - x_t = - \tilde \eta_{t + 1} \nabla f(y_t)$, hence vector $z_{t + 1} - y_t$ parallels to $x_{t + 1} - x_t$. 
                    \item Finally, with update $x_{t +1} - z_{t + 1} = L\eta_t (z_{t +1} - z_t)$, we have three colinear points $(z_t, z_{t + 1}, x_t)$ with $\Vert z_{t + 1} - z_t\Vert/\Vert z_{t +1 } - z_t\Vert = L\eta_t$. 
                \end{enumerate}
                From the above results, we can conclude that triangle $(y_t, z_t, z_{t + 1})$ is similar to $(x_t, z_t, x_{t + 1})$ because they share colinear points $z_t, y_t, x_t$ and $z_t, z_{t + 1}, x_{t + 1}$, and their sides $z_{t + 1} - y_t$ parallels to $x_{t + 1} - z_t$. 
            \end{observation}

@@

            \begin{lemma}[Similar Triangule Form I via Generic Tri Points]

@@

@@

                    z_{t + 1} &= y_t - L^{-1}\nabla f(y_t)

@@

                    x_{t + 1} &= y_t - L^{-1}\nabla f(y_t) 
                    + L \eta_t y_t - \eta_t \nabla f(y_t) - L\eta_t z_t
                    (1 + L\eta_t)y_t - (\eta_t + L^{-1})\nabla f(y_t) - L\eta_t z_t
                    &= \eta_t Lz_t + x_t -(\eta_t + L^{-1}) \nabla f(y_t)  - L\eta_t z_t
                    &= x_t - (\eta_t + L^{-1})\nabla f(y_t). 
                By assumption $\tilde \eta_{t + 1} = \eta_t + L^{-1}$, we ends with $x_{t + 1} = x_t + \tilde \eta_{t + 1}\nabla f(y_t)$. 

@@

                    z_{t + 1} &= y_t - L^{-1}\nabla f(y_t)

@@

                In this remark we clarify the name ``similar triangule" as given in the literatures. 

@@

                    The updates for $z_{t + 1}, x_{t + 1}$ are based on $y_t, x_t$ displaced by $L^{-1}\nabla f(y_t), \tilde\eta_{t +1}\nabla f(y_t)$, therefore vector $z_{t + 1} - y_t$ parallels to $x_{t + 1} - x_t$. 

@@

@@

            The next lemma will determine one choice of the stepsize sequence $\eta_t$ that recovers the classical form of Nesterov accelerated gradient method.

    \subsection{Generic Lyapunov analysis for Accelerated gradient via PPM}
        The first theorem states the generic convergence results when $f$ is differentiable with $L$-Lipschitz gradient, using \hyperref[def:agg_ppm]{definition\ref*{def:agg_ppm}}. 
        \begin{lemma}[Smooth Generic AG Lyapunov Analysis]\label{lemma:smooth_agg_lyapunov_upper_bound}
            With $f$ having $L$-Lipschitz gradient and minimizer $x_*$, using 
            \hyperref[def:agg_ppm]{definition \ref*{def:agg_ppm}} with $\mu = 0$
            and 
            \hyperref[thm:ppm_descent_ineq]{theorem \ref*{thm:ppm_descent_ineq}} ,
            then we can derive the upper bounds for the RHS of the PPM descent inequality: 
                &= \tilde \eta_{t + 1} (f(z_{t + 1}) - f(x_*))
                    \Vert x_{t + 1} - x_*\Vert^2 - \Vert x_t - x_* \Vert^2
                \frac{\tilde \eta_{t + 1} L}{2}\Vert z_{t + 1} - y_t\Vert^2
                \langle \tilde \eta_{t + 1} \nabla f(y_t), x_{t + 1} - z_{t + 1} \rangle, 
                &= f(z_{t + 1}) - f(z_t) 
                \langle \nabla f(y_t), z_{t + 1} - z_t\rangle + 
            Define $\phi_t(x) = \tilde \eta_{t +1}(f(y_t) + \langle \nabla f(y_t), x- y_t\rangle)$.  
            With $L$-smoothness of $f$ in mind, consider the following sequence of inequalities: 
                \tilde\eta_{t + 1} (f(y_t) + \langle \nabla f(y_t), x_{t + 1} - y_t\rangle)
                \phi_t (x_{t + 1}) &= \tilde \eta_{t + 1}(
            the AG generic performs PPM on $\phi_t$ to produce $x_{t + 1}$ so the PPM Lyapunov inequality from 
            \hyperref[thm:ppm_descent_ineq]{theorem \ref*{thm:ppm_descent_ineq}} applies,
            substituting yields equivalences for all $x_*$: 
                & \phi_t(x_{t + 1}) - \phi_t(x_*) + \frac{1}{2}\Vert x_* - x_{t + 1}\Vert^2 
                - \frac{1}{2} \Vert x_{t + 1} - x_t\Vert^2 
                \tilde \eta_{t + 1}\left(
                    f(z_{t + 1}) - \frac{L}{2}\Vert z_{t + 1} - y_t\Vert^2 
                    \langle \nabla f(y_t), x_{t + 1} - z_{t + 1}\rangle
                \right) - \tilde \eta_{t + 1} f(x_*)
                    \Vert x_{t + 1} - x_*\Vert^2 - \Vert x_{t} - x_*\Vert^2
                & 
                    f(z_{t + 1}) - f(x_*)
                \right) + \frac{1}{2}\Vert x_{t + 1} - x_*\Vert^2 
                - \frac{1}{2}\Vert x_{t} - x_*\Vert^2  =: \Upsilon_{1, t + 1}^{\text{AG}}
                -\frac{1}{2} \Vert x_{t + 1 } - x_t\Vert^2 + 
                \frac{\tilde \eta_{t + 1}}{2}\Vert z_{t + 1} - y_t\Vert^2 
                - \langle \tilde \eta_{t + 1}\nabla f(y_t), x_{t + 1} - z_{t + 1} \rangle. 
            Observe that, the rhs and lhs of the Lyapunov inequality are anchored at $z_{t + 1}$. 
            Similarly for the descent inequality we wish to obtain: 
                \langle \nabla f(y_t), z_{t + 1} - z_t\rangle + \frac{L}{2}\Vert z_{t + 1} - y_t\Vert^2
                \text{let }\Upsilon_{2, t + 1}^{\text{AG}} 
                &:= f(z_{t + 1}) - f(z_t). 
            Which is the descent inequality anchored on $z_{t + 1}$.
            Merging the $(z_{t + 1} - y_t)$ with $y_t - z_t$ together yield the desired results. 

@@

    \subsection{Scenario I}

@@

    \subsection{Scenario II}
        With the choice of $\tilde \eta_{t + 1} = \eta_t + L^{-1}$, we can recover the seminal Nesterov accelerated gradient algorithm from 1983. 
        The variants proposed by Chambolle Dossal \cite{chambolle_convergence_2015}, is distinct from the possibility proposed in this secenario. 
        % Recall similar a lemma about similar triangle form here.

@@

            allows \hyperref[def:agg_tri]{definition \ref*{def:agg_tri}} to converge at an optimal rate. 

@@

            Continue from \ref*{lemma:smooth_agg_lyapunov_upper_bound}, With $x_{t +1} - x_t = L\tilde\eta_{t+1}(z_{t + 1} - y_t)$ for $y_t$: 

@@

                - \frac{1}{2}\Vert x_{t+1 } - x_t\Vert^2 + 

@@

                    \tilde \eta_{t+ 1} \nabla f(y_t), x_{t+1} - z_{t+1}

@@

                \Upsilon_{2, t + 1}^{\text{AG}} &= 
                \langle \nabla f(y_t), z_{t+1} - y_t\rangle
                \frac{L }{2} \Vert z_{t+1} - y_t\Vert^2 + 
                \langle \nabla f(y_t), z_{t + 1} - z_t\rangle

@@

@@

            If the above quantity less than or equal to zero, it will allow for convergence of the algorithm. 

@@

            Combining yeilds that 

@@

            We had recovered the classic Nesterov accelerated sequence method for the smooth gradient descent algorithm. 

@@

            It fails the similar triangule description Without the explict representation $x_{n + 1} = z_n + \tilde\eta_{t + 1}\nabla f(y_t)$. 

>>>>

    \par\noindent
    \textbf{We state our contributions}. 
    In \hyperref[sec:AG_varieties]{section \ref*{sec:AG_varieties}} 
    We stated different forms of accelerated gradient that appeared in the literatures and the equivalences between their forms. 
    They are abstract because the choice of stepsize parameters in the algorithms are unspecified. 
    Abstract forms related to the proximal gradient method were our own inventions inspired by discussions in the literatures, which includes 
    \hyperref[def:ag_prox_grad_ppm]{definition \ref*{def:ag_prox_grad_ppm}}, and 
    \hyperref[def:ag_prox_grad_tri_pt]{definition \ref*{def:ag_prox_grad_tri_pt}}. 
    In \hyperref[sec:AG_useful_information]{section \ref*{sec:AG_useful_information}}
    we prove and interpret some of the unproved claims made by Ahn, Sra \cite{ahn_understanding_2022} as good exercises and for the peace of the mind. 
    In
    \hyperref[sec:generic_ag_ppm_lyapunov_analysis]{section \ref*{sec:generic_ag_ppm_lyapunov_analysis}},
    we state two important theorems for the derivation of convergences rate for various type of algorithm using an upper bound of a Lyapunov function. 
    This part is inspired by works from Ahn, Sra \cite{ahn_understanding_2022}, but with the gradient mapping and proximal gradient added into the picture, making it distinct from all algorithms discussed in Ahn and Sra's paper. 
    \hyperref[sec:scenario_i]{Section \ref*{sec:scenario_i}} and 
    \hyperref[sec:scenario_ii]{\ref*{sec:scenario_ii}} derive the convergence rate and determine the parameters for the step sizes needed to achieve an optimal convergence rate. 
    In each section, we recover a specific variant of the Accelerated gradient and identify their appearances in the literature.

@@

            \begin{definition}[Similar Triangle Form I]
            \label{def:ag_form_similar_tria_I}
                It updates iterates $(y_t, x_{t + 1}, z_{t + 1})$ using 
                \begin{align*}
                    z_{t + 1} &= y_t - L^{-1} \mathcal G_L(y_t) 
                    \\
                    x_{t + 1} &= z_{t + 1} + L\eta_t (z_{t + 1} - z_t)
                    \\
                    y_{t + 1} &= 
                    (1 + L\eta_{t + 1})^{-1}
                    (
                    x_{t + 1} + L\eta_{t + 1}z_{t + 1}
                    ). 
                \end{align*}
            \end{definition}
            \begin{observation}
                There are a lot of observations about this form to unpack. 
                Let's list the following observations
                \begin{enumerate}
                    \item Observe that $y_t = (1 + L\eta_t)^{-1}(x_t + L\eta_t z_t)$ is a convex combination between $x_t, z_t$. Hence $z_t, y_t, x_t$ are three collinear points. Additionally, we have $y_t - x_t = L\eta_t (z_t - y_t)$, hence $\Vert y_t - x_t\Vert/\Vert z_t - y_t\Vert = L\eta_t$. 
                    \item By the algorithm, we have $z_{t + 1} - y_t = - L^{-1} \nabla f(y_t)$, and $x_{t + 1} - x_t = - \tilde \eta_{t + 1} \nabla f(y_t)$, hence vector $z_{t + 1} - y_t$ parallels to $x_{t + 1} - x_t$. 
                    \item Finally, with update $x_{t +1} - z_{t + 1} = L\eta_t (z_{t +1} - z_t)$, we have three colinear points $(z_t, z_{t + 1}, x_t)$ with $\Vert z_{t + 1} - z_t\Vert/\Vert z_{t +1 } - z_t\Vert = L\eta_t$. 
                \end{enumerate}
                From the above results, we can conclude that triangle $(y_t, z_t, z_{t + 1})$ is similar to $(x_t, z_t, x_{t + 1})$ because they share colinear points $z_t, y_t, x_t$ and $z_t, z_{t + 1}, x_{t + 1}$, and their sides $z_{t + 1} - y_t$ parallels to $x_{t + 1} - z_t$. 
            \end{observation}
            \begin{remark}
                For more information about the specific choice of $\eta_t$ allows for an optimal convergence rates, see 
                \hyperref[sec:scenario_ii]{section \ref*{sec:scenario_ii}} for more information. 
            \end{remark}

@@

            \begin{lemma}[Similar Triangle Form I via Generic Tri Points]

@@

                    z_{t + 1} &= y_t - L^{-1} \mathcal G_L(y_t)

@@

                    z_{t + 1} &= y_t - L^{-1}\mathcal G_L(y_t)

@@

                    x_{t + 1} &= y_t - L^{-1}\mathcal G_L(y_t) 
                    + L \eta_t y_t - \eta_t \mathcal G_L(y_t) - L\eta_t z_t
                    (1 + L\eta_t)y_t - (\eta_t + L^{-1})\mathcal G_L(y_t) - L\eta_t z_t
                    &= \eta_t Lz_t + x_t -(\eta_t + L^{-1}) \mathcal G_L(y_t)  - L\eta_t z_t
                    &= x_t - (\eta_t + L^{-1})\mathcal G_L(y_t). 
                So $x_{t + 1} = x_t + \tilde \eta_{t + 1}\mathcal G_L(y_t)$, by assumption $\tilde \eta_{t + 1} = \eta_t + L^{-1}$

@@

                    z_{t + 1} &= y_t - L^{-1} \mathcal G_L f(y_t)

@@

                In this remark we clarify the name ``similar triangle" as given in the literatures. 

@@

                    The updates for $z_{t + 1}, x_{t + 1}$ are based on $y_t, x_t$ displaced by $L^{-1} \mathcal G_L(y_t), \tilde\eta_{t +1} \mathcal G_L(y_t)$, therefore vector $z_{t + 1} - y_t$ parallels to $x_{t + 1} - x_t$. 

@@

                Finally, similar remarks about the similar triangle form can be found in Ahn, Sra's paper \cite{ahn_understanding_2022} as well. 

@@

    \subsection{Generic Lyapunov analysis for Accelerated gradient via PPM}
    \label{sec:generic_ag_ppm_lyapunov_analysis}
        The first theorem states the generic convergence results when $f$ is differentiable with $L$-Lipschitz gradient, using \hyperref[def:agg_ppm]{definition\ref*{def:agg_ppm}}.

@@

    \subsection{Scenario I}\label{sec:scenario_i}

@@

    \subsection{Scenario II}\label{sec:scenario_ii}
        With the choice of $\tilde \eta_{t + 1} = \eta_t + L^{-1}$, we can recover the seminal Nesterov accelerated gradient algorithm from 1983. 
        The variants proposed by Chambolle Dossal \cite{chambolle_convergence_2015}, is distinct from the possibility proposed in this secenario.

@@

            allows \hyperref[def:ag_prox_grad_tri_pt]{definition \ref*{def:ag_prox_grad_tri_pt}} to converge at an optimal rate. 

@@

            \hyperref[lemma:nsmooth_agg_lyapunov_upper_bound]{lemma \ref*{lemma:nsmooth_agg_lyapunov_upper_bound}}, 
            With $x_{t +1} - x_t = L\tilde\eta_{t+1}(z_{t + 1} - y_t)$ for $y_t$: 

@@

@@

                        \tilde \eta_{t+ 1} \mathcal G_L(y_t), x_{t+1} - z_{t+1}

@@

                \langle \mathcal G_L(y_t), z_{t + 1} - z_t\rangle

@@

                \langle \mathcal G_L(y_t), L^{-1}\eta_t^{-1}(x_{t+1} - z_{t + 1})\rangle

@@

                -\tilde\eta_{t + 1}^2 + \tilde\eta_{t + 1}
                    + \sum_{i = 1}^{t}\tilde\eta_i
                \right)\Vert z_{t + 1} - y_t\Vert^2
                    L^{-1}\eta_t^{-1} 
                \right)\langle \mathcal G_L (y_t), x_{t + 1} - z_{t + 1}\rangle
            If the above quantity less than or equal to zero, it will allow for convergence of the algorithm by 
            \hyperref[thm:generic_ag_convergence]{theorem \ref*{thm:generic_ag_convergence}}.

@@

            Combining yields that 

@@

            We had recovered the classic Nesterov acceleration sequence method for the smooth gradient descent algorithm. 

@@

            It fails the similar triangle description Without the explict representation $x_{n + 1} = z_n + \tilde\eta_{t + 1}\nabla f(y_t)$. 
